---
title: "March Madness Analysis"
author: "Loren Lipsey"
date: "March 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 99)
```

# March Madness Expected Value calcualation
### For seed-weighted scoring, with Nate Silver's predictions

Computes net expected value, based on the following scoring system:

+--------+-----------+
| Round  |  Score    |
+========+===========+
|First   |  1 + seed |
+--------+-----------+
|Second  |  2 + seed |
+--------+-----------+
|Sweet 16|  4 + seed |
+--------+-----------+
|Elite 8 |  8 + seed |
+--------+-----------+
|Final 4 | 16 + seed |
+--------+-----------+
|Champ   | 32 + seed |
+--------+-----------+


Download data from Five Thirty Eight:
```{r data, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)

NCAA_data <- read_csv("https://projects.fivethirtyeight.com/march-madness-api/2017/fivethirtyeight_ncaa_forecasts.csv")

```

Remove letters from seed variable:
```{r}
NCAA_data$team_seed <- gsub("[^0-9]", "", NCAA_data$team_seed)
NCAA_data$team_seed <- as.numeric(NCAA_data$team_seed)
```


Calculate the Expected Values for each team in each round, as:

## **Expected value for round = **
## _(Cumulative value of round-dependent points_
## _+ cumulative value of low-seed bonus points)_
## _* chance of getting this far and winning_
```{r}
NCAA_EV <- NCAA_data %>%  
  mutate(EV_second = (1 + team_seed) * rd2_win) %>%
  mutate(EV_sweet16 = (3 + (2 * team_seed)) * rd3_win) %>%
  mutate(EV_elite8 = (7 + (3 * team_seed)) * rd4_win) %>%
  mutate(EV_final4 = (15 + (4 * team_seed)) * rd5_win) %>%
  mutate(EV_champgame = (31 + (5 * team_seed)) * rd6_win) %>%
  mutate(EV_winner = (63 + (6 * team_seed)) * rd7_win)  %>%
  select(contains("EV"), team_seed, team_region, team_name) 
```


Merge EVs for teams in the play-in games:
```{r}
NCAA_EV <- NCAA_EV %>%
  group_by(team_seed, team_region) %>%
  mutate(
    EV_second = sum(EV_second),
    EV_sweet16 = sum(EV_sweet16),
    EV_elite8 = sum(EV_elite8),
    EV_final4 = sum(EV_final4),
    EV_champgame = sum(EV_champgame),
    EV_winner = sum(EV_winner)
  ) %>%
  ungroup
```


## Champ round top picks
```{r}
NCAA_EV %>%
  select(team_name, EV_winner, EV_champgame) %>%
  arrange(desc(EV_winner)) %>%
  head(15) %>%
  kable("html", digits = 2)
```



## West Region
```{r}
NCAA_EV %>%
  select(team_name, team_region, EV_second, EV_sweet16, EV_elite8, EV_final4) %>%
  filter(team_region == "West") %>%
  arrange(desc(EV_final4)) %>%
  kable("html", digits = 2)
```



## East Region
```{r}
NCAA_EV %>%
  select(team_name, team_region, EV_second, EV_sweet16, EV_elite8, EV_final4) %>%
  filter(team_region == "East") %>%
  arrange(desc(EV_final4)) %>%
  kable("html", digits = 2)
```


## Midwest Region
```{r}
NCAA_EV %>%
  select(team_name, team_region, EV_second, EV_sweet16, EV_elite8, EV_final4) %>%
  filter(team_region == "Midwest") %>%
  arrange(desc(EV_final4)) %>%
  kable("html", digits = 2)
```


## South Region
```{r}
NCAA_EV %>%
  select(team_name, team_region, EV_second, EV_sweet16, EV_elite8, EV_final4) %>%
  filter(team_region == "South") %>%
  arrange(desc(EV_final4)) %>%
  kable("html", digits = 2)
```